Genome-wide changes in expression during inflorescence architecture development

### 2015-10-27 ms.Rmd
Added intro

## Rice inflorescence development
## Branching & grain yield
## Sequential changes in meristem activity
## Discuss a few key genes?

@@

## Dynamic expression of homeodomain genes

@@

>>>>

Agricultural development is essential to ensure food production and security for a growing population, particularly in developing countries [@Borlaug:2007cp]. Rice is a staple food for over half the world's population, including a large number of developing countries. A sustainable increase in the production of rice against a backdrop of climate change and diminishing water and land availability will require plants with improved grain output, and the establishment of high-yield varieties is a major goal of modern breeding programs [@Peng:2008jb]. Rice yield is a complex trait influenced by genetic factors, as well as environmental and other epigenetic factors, and is progressively defined during the life cycle of the plant, first during the vegetative phase, where the number of fertile tillers is established, and then during the reproductive phase and grain-filling phase. Yield is directly dependent on the architecture of the inflorescence [@Ikeda:2004tk], which is the flower-bearing structure. The rice inflorescence (or panicle) is a branched structure composed of a rachis (the main axis), primary branches, higher order branches, and spikelets. Spikelets are short branches that form on other branches of the panicle and bear a single flower. They are the primary unit of grass inflorescence, with one spikelet bearing one floret, meaning that grain yield is determined by the number of spikelets [@Ikeda:2004tk; @Yoshida:2011bo].

During panicle development, the undifferentiated cells in the primary axis (i.e. the rachis meristem, RM) undergo transitions through a series of meristem identities. After the establishment of the rachis some RM cells differentiate into branch meristem (BM) **PBM/SBM** for the formation and elongation of primary branches and the establishment of higher-order branches. The axillary BM cells differentiate into spikelet meristems (SM), which develop into floret meristems (FM) and produce flowers. After spikelet differentiation, panicle branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk].

The landscape of gene expression is a characteristic factor that differentiates meristem cell types, and is involved in the control of transitions between meristem identities and therefore panicle phenotype. Mutant analyses and map-based cloning of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq]. Some of these genes are involved in the patterning of axillary meristem and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq]. A large set of genes related to floral development, which may also affect panicle architecture, has been identified [@Yoshida:2011bo], and some of them may have been under selection during domestication or are associated with crop improvement in Asian rice  [@Xing:2010de; @He:2011io; @Wang:2011kq; @Xu:2012hq; @Ikeda:2013go].

Understanding of the events that occur during development that determine panicle characteristics such as branching complexity and its plasticity will be vital for sustainable improvement of rice yield potential using targeted breeding programs. Despite advances in characterization at the level of individual genes or interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and the environment [@Azpeitia:2013jia]. One step towards this goal is to describe the changes in gene expression that occur during development. In this report, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of rice panicle development, using a sensitive LASER microdissection and RNA sequencing approach.

@@

## Patterns of expression of homeodomain genes

@@

## Summary

This report describes the changes in gene expression that occur during the early stages of rice panicle development. A switch in expression between apical and axillary meristems followed by gradual changes between PBM and SM were observed, based on changes in expression of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling. These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.
